Relict deciduous oaks show high resilience to drought along a climatic gradient in the rear-edge despite strong land-use legacies 

### 2019-08-05 man/ecosystems/ms_ecosystems.Rmd
update Results

The analysis of greenness time trends showed that EVI in 78.9 % of the pixels of *Quercus pyrenaica* forests experienced a positive trend for the 2000-2016 period. The lowest values of EVI standardized anomalies for the studied period were recorded during the 2005 drought, and the minimum EVI values were expressed in northern (dry) populations (Figure 2). In fact, a "browning" episode ($\mathrm{EVI_{sa}} < -1$) was observed for *Q. pyrenaica* forests during this drought event, whereas no changes in greenness were observed in response to the 2012 drought (Figure 3).

@@

The trees of the southern population were older than those from the northern one. In addition trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites (Table 1, Figure 4). Both growth and height of trees from the northern and the low-elevation southern population were similar (Figures 4 and S6). Competition was similar among all sites. Yet plot basal area was greatest in CA-High (Table 1), despite the highest individual BAI growth at that site (Figure 4). Only trees from the southern sites showed significant positive growth trends since the late 1970s (Figure 4), being more pronunced for the highest site (CA-High).

Drought events produced reduction of RWI for all sites (Figure S3), being particularly evident the negative effect of the 1995 drought, when the greatest reduction of tree-growth were recorded for all chronologies. The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S3). Tree-growth reduction followed a moisture-gradient, being weaker for the wetter site (CA-High). Trees from northern site also suffered a great reduction during 1945-1946 drought event (Figure 5).

The northern site (SJ) showed two major release events (GC > 50 % occurring in more than 50 % of sampled trees): the first during the 1940 decade (the most evindent) and the second in the period 1995-2000. These periods alternate with periods of supression for over 40 years. The two southern sites showed no release events except for CA-High at the beginning of the 1830s and no suppression events in the last 50 years.

## Resilience to drought events at the ecosystem and individual tree levels
The exploration of the resilience metrics of tree-growth for severe drought events since 1950 (*i.e.* for the share period among the three chronologies, excluding juvenile years, Table S3) revealed a positive relation between drought severity and Recovery (*Rc*), significant for all oak populations (Figure 6). A similar pattern was obtained for Resilience (*Rs*) but only significant for SJ site (Figure 6). Non-significant patterns were obtained if we exclude 1995 from the analysis (Figure S8). Trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area (Table S3), and particularly SJ, where our results suggest a major release event after 1995 (Figure 5).

During the last two drought events, resilience metrics for greenness and tree-growth were significantly different bewteen drought events (Table 3). The 2005 drought event reduced greennes and growth more than that of 2012 (Tables S1 and S2) but the two metrics of resilience generally covaried in the same direction. We obtained significantly higher Resilience (*Rs*) and Resistance (*Rt*) values for the 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 7). Recovery (*Rc*) showed contrasting pattern for EVI and tree-growth. Recovery values for greenness was higher for 2005 than for 2012 drought event (Table S1, Figure 7).

Resilience metrics significantly varied between sites, except for Resilience (*Rs*) of tree-growth (p = 0.534; Table 3). The two southern populations showed lower values for Recovery (*Rc*) than the northern site both for greenness and tree-growth (Tables S1 and S2). In contrast, Resistance and Resilience values were significantly higher for southern sites than for northern ones (Tables S1 and S2).

## Tree-growth response to climate
Tree-growth exhibited a significant greater response to water availability than to temperatures, as it can be expected for a Mediterranean site and a species located closer to its xeric/dry distribution limit (rear-edge). Cumulative precipitation of the hydrological year was the climatic variable with the highest relationship with growth and influenced similarly and positively the tree-growth for both northern and southern populations (Figure 8a). Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population. Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 8a), being this relationship higher in the northern population (r > 0.6 *vs.* r < 0.5). Relationships with temperatures were lower than those with moisture variables. Yet there were differences between northern and southern populations: spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while no significant limitations were recorded for northen site regarding the temperatures (Figure 8b and 8c). Remarkably minimum and maximum temperatures of current September positively influenced the tree-growth only in the northern population. No positive relationship with temperature was found for southern trees.

>>>>

The analysis of greenness time trends showed that EVI in 78.9 % of the pixels of *Quercus pyrenaica* forests experienced a positive trend for the 2000-2016 period. The lowest values of EVI standardized anomalies for the studied period were recorded during the 2005 drought, and the minimum EVI values were expressed in northern (dry) populations (Figure 2). A "browning" episode ($\mathrm{EVI_{sa}} < -1$) was observed during this drought event, whereas no changes in greenness were observed in response to the 2012 drought (Figure 3).

@@

The trees of the southern population were older than those from the northern one. In addition, trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites (Table 1, Figure 4), despite competition measured as plot basal area was greatest in CA-High (Table 1). Growth and height of trees from the northern and the low-elevation southern population were similar (Figures 4 and S6). Only trees from the southern sites showed significant positive growth trends since the late 1970s (Figure 4), being much more pronounced for the highest, hence more humid, site (CA-High).

Drought events produced reduction in RWI for all sites (Figure S3), being particularly evident the negative effect of the 1995 drought (the worst drought spell in our climatic record, Table S3), when the greatest growth reduction was recorded for all chronologies. The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S3). Tree-growth reduction followed a moisture-gradient, being weaker for the wettest site (CA-High, Figure 5).

The northern site (SJ) showed two major release events (GC > 50 % occurring in more than 50 % of sampled trees): the first during the 1940s (the most evident) and the second in 1995-2000. These periods alternated with periods of suppression. In contrast, the two southern sites showed no release events except for CA-High at the beginning of the 1830s and no suppression events in the last 50 years.

## Resilience to drought events at the ecosystem and individual tree levels
Resilience metrics of tree-growth for drought events since 1950 (*i.e.* shared period among the three chronologies excluding the juvenile years, Table S3) revealed a positive relationship between drought severity and Recovery (*Rc*), significant for all oak populations (Figure 6). A similar pattern was obtained for Resilience (*Rs*) but only significant for SJ site (Figure 6). Importantly, non-significant patterns were obtained if we excluded 1995 from the analysis, except for *Rc* in SJ (Figure S8). Trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area (Table S3), and particularly SJ, where our results suggest a major release event after 1995 (Figure 5).

During the last two drought events, resilience metrics for greenness and tree-growth were significantly different between drought events (Table 3). The 2005 drought event reduced greenness and growth more than that of 2012 (Tables S1 and S2) but the two metrics of resilience generally covaried in the same direction. We obtained significantly higher Resilience (*Rs*) and Resistance (*Rt*) values for the 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 7). Recovery (*Rc*) showed contrasting pattern for EVI and tree-growth. Recovery values for greenness were higher for 2005 than for 2012 drought event (Table S1, Figure 7).

Overall, resilience metrics significantly varied between sites, except for Resilience (*Rs*) of tree-growth (p = 0.534; Table 3), which was similar among the three sites. The two southern populations showed lower values for Recovery (*Rc*) than the northern site both for greenness and tree-growth (Tables S1 and S2). In contrast, Resistance and Resilience values were significantly higher for southern sites than for northern ones (Tables S1 and S2).

## Tree-growth response to climate
The response of tree-growth to water availability was greater than that to temperatures, as it can be expected for a Mediterranean site and particularly a species located closer to its xeric/dry distribution limit (rear-edge). Cumulative precipitation of the hydrological year and seasonal SPEI values (*i.e.* for the Hydrological year, Spring and Summer) were the climatic variables exhibiting the highest (positive) relationship with growth for all populations (Figure 8a). Yet there were sone differences between northern and southern populations: the positive relationship with SPEI was highest in the northern population (r > 0.6 *vs.* r < 0.5; Figure 8a). In addition, spring maximum temperature was the most significant limitation for tree-growth only for the southern populations (Figure 8b and 8c). Minimum and maximum temperatures of current September positively influenced tree-growth only in the northern population.
